News Across Languages - Cross-Lingual Document Similarity and Event Tracking

### 2016-01-06 rupnik2015.tex
fixed bib and style changes

\jairheading{x}{xxxx}{x-xx}{x/xx}{x/xx}
\ShortHeadings{Cross-Lingual Document Similarity and Event Tracking}
{Rupnik, Muhi\v{c}, Leban, Fortuna, \v{S}kraba \& Grobelnik}
\firstpageno{1}

@@

In today's world, we follow news which is distributed globally. Significant events are reported by different sources and in different languages. In this work, we address the problem of tracking of events in a large multilingual stream. Within a recently developed system Event Registry~\shortcite{Leban2014W,Leban2014I} we examine two aspects of this problem: how to compare articles in different languages and how to link collections of articles in different languages which refer to the same event.  Taking a multilingual stream and clusters of articles from each language, we compare different cross-lingual document similarity measures based on Wikipedia. This allows us to compute the similarity of any two articles regardless of language. Building on previous work, we show there are methods which scale well and can compute a meaningful similarity between articles from languages with little or no direct overlap in the training data.
Using this capability, we then propose an approach to link clusters of articles across languages which represent the same event. We provide an extensive evaluation of the system as a whole, as well as an evaluation of the quality and robustness of the similarity measure and the linking algorithm.
\end{abstract}

@@

Content on the Internet is becoming increasingly multilingual. A prime example is Wikipedia. In 2001, the majority of pages were written in English, while in 2015, the percentage of English articles has dropped to 14\%. At the same time, online news has begun to dominate reporting of current events. However, machine translation remains relatively rudimentary. It allows people to understand simple phrases on web pages, but remains inadequate for more advanced understanding of text. In this paper we consider the intersection of these developments: how to track events which are reported about in multiple languages.

@@

Collected articles are first semantically annotated by identifying mentions of relevant concepts -- either entities or important keywords. The disambiguation and entity linking of the concepts is done using Wikipedia as the main knowledge base. The algorithm for semantic annotation uses machine learning to detect significant terms within unstructured text and link them to the appropriate Wikipedia articles. The approach models link probability and combines prior word sense distributions with context based sense distributions. The details are reported in~\cite{Milne:2008:LLW:1458082.1458150} and ~\cite{zhang2014saaacamactat}. As a part of the semantic annotation we also analyze the dateline of the article to identify the location of the described event as well as to extract dates mentioned in the article using a set of regular expressions. Since articles are frequently revised we also detect if the collected article is just a revision of a previous one so that we can use this information in next phases of the pipeline. The last important processing step on the document level is to efficiently identify which articles in other available languages are most similar to this article. The methodology for this task is one of the main contributions of this paper and is explained in details in Section~\ref{sec:crosslingual}.

As the next step, an online clustering algorithm is applied to the articles in order to identify groups of articles that are discussing the same event. For each new article, the clustering algorithm determines if the article should be assigned to some existing cluster or into a new cluster. The underlying assumption is that articles that are describing the same event are similar enough and will therefore be put into the same cluster. For clustering, each new article is first tokenized, stop words are removed and the remaining words are stemmed.
The remaining tokens are represented in a vector-space model and normalized using TF-IDF\footnote{The IDF weights are dynamically computed for each new article over all news articles within a 10 day window.} (see Section~\ref{sec:tfidf} for the definition). Cosine similarity is used to find the most similar existing cluster, by comparing the document's vector to the centroid vector of each cluster. A user-defined threshold is used to determine if the article is not similar enough to any existing clusters (0.4 was used in our experiments).  If the highest similarity is above the threshold, the article is assigned to the corresponding cluster, otherwise a new cluster is created, initially containing only the single article. Whenever an article is assigned to a cluster, the cluster's centroid vector is also updated. Since articles about an event are commonly written only for a short period of time, we remove clusters once the oldest article in the cluster becomes more than 4 days old. This housekeeping mechanism prevents the clustering from becoming slow and also ensures that articles are not assigned to obsolete clusters. Details of the clustering approach are described in~\cite{brank2014}.

@@

When a cluster is identified and information about the event is extracted, all available data is stored in a custom-built database system. The data is then accessible through the API or a web interface\footnote{http://eventregistry.org/}, which provide numerous search and visualization options.

@@

\section{Related work}\label{sec:related}

@@

\subsection{Cross-lingual document similarity}

@@

\textbf{Translation and dictionary based}. The most obvious way to compare documents written in different languages is to use machine translation and perform monolingual similarity, see  ~\cite{multilingualBook,plagiarism} for several variations of translation based approaches. One can use free tools such as Moses~\shortcite{moses} or translation services, such as Google Translate\footnote{https://translate.google.com/}. There are two issues with such approaches: they solve a harder problem than needs to be solved and they are less robust to training resource quality - large sets of translated sentences are typically needed. Training Moses for languages with scarce linguistic resources is thus problematic. The issue with using online services such as Google Translate is that the APIs are limited and not free. The operation efficiency and cost requirements make translation-based approaches less suited for our system. Closely related are works Cross-Lingual Vector Space Model (CL-VSM)~\cite{plagiarism} and the approach presented in~\cite{pouliquen2008story} which both compare documents by using dictionaries, which in both cases are EuroVoc dictionaries~\cite{eurovoc}. The generality of such approaches is limited by the quality of available linguistic resources, which may be scarce or non-existent for certain language pairs.

\textbf{Probabilistic topic models}. There exist many variants to modelling documents in a language independent way by using probabilistic graphical models. The models include:  Joint Probabilistic Latent Semantic Analysis (JPLSA)~\shortcite{platt2010translingual}, Coupled Probabilistic LSA (CPLSA)~\cite{platt2010translingual}, Probabilistic Cross-Lingual LSA (PCLLSA)~\cite{PCL_LSA} and Polylingual Topic Models (PLTM)~\shortcite{polyLDA} which is a Bayesian version of PCLLSA. The methods (except for CPLSA) describe the multilingual document collections as samples from generative probabilistic models, with variations on the assumptions on the model structure. The topics represent latent variables that are used to generate observed variables (words), a process specific to each language. The parameter estimation is posed as an inference problem which is typically intractable and one usually solves it using approximate techniques. Most variants of solutions are based on Gibbs sampling or Variational Inference, which are nontrivial to implement and may require an experienced practitioner to be applied. Furthermore, representing a new document as a mixture of topics is another potentially hard inference problem which must be solved.

\textbf{Matrix factorization}. Several matrix factorization based approaches exist in the literature. The models include: Non-negative matrix factorization based~\cite{nonnegfactor_lsi}, Cross-Lingual Latent Semantic Indexing CL-LSI~\shortcite{cl_lsi,multilingualBook}, Canonical Correlation Analysis (CCA)~\cite{Hotelling}, Oriented Principal Component Analysis (OPCA)~\cite{platt2010translingual}. The quadratic time and space dependency of the OPCA method makes it impractical for large scale purposes. In addition, OPCA forces the vocabulary sizes for all languages to be the same, which is less intuitive. For our setting, the method in ~\cite{nonnegfactor_lsi} has a prohibitively high computational cost when building models (it uses dense matrices whose dimensions are a product of the training set size and the vocabulary size). Our proposed approach combines CCA and CL-LSI. Another closely related method is Cross-Lingual Explicit Semantic Analysis (CL-ESA)~\cite{ESA}, which uses Wikipedia (as do we in the current work) to compare documents. It can be interpreted as using the sample covariance matrix between features of two languages to define the dot product which is used to compute similarities.
The authors of CL-ESA compare it to CL-LSI and find that CL-LSI can outperform CL-ESA in an information retrieval, but is costlier to optimize over a large corpus (CL-ESA requires no training). We find that the scalability argument does not apply in our case: based on advances in numerical linear algebra we can solve large CL-LSI problems (millions of documents as opposed to the 10,000 document limit reported in~\cite{ESA}). In addition, CL-ESA is less suited for computing similarities between two large monolingual streams. For example, each day we have to compute similarities between 500,000 English and 500,000 German news articles. Comparing each German news article with 500,000 English news articles is either prohibitively slow (involves projecting all English articles on Wikipedia) or consumes too much memory (involves storing the projected English articles, which for a Wikipedia of size 1,000,000 is a 500,000 by 1000,0000 non-sparse matrix).

\textbf{Monolingual}. Finally, related work includes monolingual approaches that treat document written in different languages in a monolingual fashion. The intuition is that named entities (for example, ``Obama'') and cognate words (for example, ``tsunami'') are written in the same or similar fashion in many languages. For example, the Cross-Language Character n-Gram Model (CL-CNG)~\cite{plagiarism} represents documents as bags of character $n$-grams. Another approach is to use language dependent keyword lists based on cognate words ~\cite{pouliquen2008story}. These approaches may be suitable for comparing documents written in languages that share a writing system, which does not apply to the case of global news tracking.

@@

\subsection{Cross-lingual cluster linking}

@@

A system, which is significantly different but worth mentioning, is the GDELT project~\cite{Leetaru2013Gdelt}. In GDELT, events are also extracted from articles, but in their case, an event is specified in a form of a triple containing two actors and a relation. The project contains an extensive vocabulary of possible relations, mostly related to political events. In order to identify events, GDELT collects articles in more than 65 languages and uses machine translation to translate them to English. All information extraction is then done on the translated article.

\section{Cross-lingual Document Similarity}
\label{sec:crosslingual}
Document similarity is an important component in techniques from text mining and natural language processing. Many techniques use the similarity as a black box, e.g., a kernel in Support Vector Machines. Comparison of documents (or other types of text snippets) in a monolingual setting is a well-studied problem in the field of information retrieval ~\cite{Salton88term-weightingapproaches}. We first formally introduce the problem followed by a description of  our approach.

\subsection{Problem definition}\label{sec:tfidf}
We will first describe how documents are represented as vectors and how to compare documents in a mono-lingual setting. We then define a way to measure cross-lingual similarity which is natural for the models we consider.

\noindent\textbf{Document representation.}
The standard vector space model~\cite{Salton88term-weightingapproaches} represents documents as vectors, where each term corresponds to a word or a phrase in a fixed vocabulary. Formally, document $d$ is represented by a vector $x \in \RR^n$, where $n$ corresponds to the size of the vocabulary, and vector elements $x_k$ correspond to the number of times term $k$ occurred in the document, also called \emph{term frequency} or $TF_k(d)$.

@@

\noindent\textbf {Mono-lingual similarity.}
A common way of computing similarity between documents is \emph{cosine similarity},
$$sim(d_1, d_2) = \frac{\langle \phi(d_1), \phi(d_2)\rangle}{\|\phi(d_1)\| \|\phi(d_2)\|},$$
where $\langle \cdot,\cdot \rangle$ and $\|\cdot\|$ are standard inner product and Euclidean norm. When dealing with two or more languages, one could ignore the language information
and build a vector space using the union of tokens over the languages. A cosine similarity function in such a space can be useful to some extent, for example ``Internet'' or ``Obama'' may appear both in Spanish and English texts and the presence of such terms in both an English and a Spanish document would contribute to their similarity. In general however, large parts of vocabularies may not intersect. This means that given a language pair, many words in both languages cannot contribute to the similarity score. Such cases can make the similarity function very insensitive to the data.

\noindent\textbf {Cross-lingual similarity.}
Processing a multilingual dataset results in several vector spaces with varying dimensionality, one for each language. The dimensionality of the vector space corresponding to the $i$-th language is denoted by $n_i$ and the vector space model mapping is denoted by $\phi_i : \text{text} \rightarrow \RR^{n_i}$.
The similarity between documents in language $i$ and language $j$ is defined as a bilinear operator represented as a matrix $S_{i,j} \in \RR^{n_i \times n_j}$:
$$sim_{i,j}(d_1, d_2) = \frac{ \langle \phi_i (d_1), S_{i,j} \phi_j (d_2) \rangle }{\|\phi_i(d_1)\| \|\phi_j(d_2)\|},$$
where $d_1$ and $d_2$ are documents written in the $i$-th and $j$-th language respectively. If the maximal singular value of $S_{i,j}$ is bounded by $1$, then the similarity scores will lie on the interval $[-1, 1]$. We will provide an overview of the models in Section \ref{sec:models} and then introduce additional notation in \ref{sec:notation}. Starting with Section \ref{sec:kmeans} and ending with Section \ref{sec:hublang} we will describe some approaches to compute $S_{i,j}$ given training data.

\subsection{Cross-Lingual Models}\label{sec:models}
In this section, we will describe several approaches to the problem of computing the multilingual similarities introduced in Section~\ref{sec:tfidf}. We present four approaches:
a simple approach based on $k$-means clustering in Section~\ref{sec:kmeans}, a standard approach based on singular value decomposition in Section~\ref{sec:LSI}, a related
approach called Canonical Correlation Analysis (CCA) in Section~\ref{sec:CCA} and finally a new method, which is an extension of CCA to more than two languages in Section~\ref{sec:hublang}.
%
CCA can be used to find correlated patterns for a pair of languages, whereas the extended method optimizes a
Sum of Squared Correlations (SSCOR) between several language pairs, which was introduced in~\cite{Kettenring}. The SSCOR problem is difficult to solve in our setting (hundreds of thousands of features, hundreds of thousands of examples). To tackle this, we propose a method which consists of two ingredients.
 The first one is based on an observation that certain datasets (such as Wikipedia) are biased towards one language (English for Wikipedia), which can be exploited
 to reformulate a difficult optimization problem as an eigenvector problem. The second ingredient is dimensionality reduction using CL-LSI, which
 makes the eigenvector problem computationally and numerically tractable.

@@

We will now describe four models to cross-lingual similarity computation in the next sub-sections.
\subsection{$k$-means}\label{sec:kmeans}

@@

\paragraph{Implementation note}

@@

\subsection{Canonical Correlation Analysis}\label{sec:CCA}
 We now present a statistical technique to analyze data from two sources, an extension of which will be presented in the next section.
%
 Canonical Correlation Analysis (CCA)~\cite{Hotelling} is a dimensionality reduction technique similar to Principal Component Analysis (PCA)~\cite{Pearson1901On}, with the additional assumption that the data consists of feature vectors that arose from two sources (two views) that share some information. Examples include: bilingual document collection ~\shortcite{mrpqr} and collection of images and captions~\shortcite{Hardoon_usingimage}. Instead of looking for linear combinations of features that maximize the variance (PCA) we look for a linear combination of feature vectors from the first view and a linear combination for the second view, that are maximally correlated.

Interpreting the columns of $X_i$ as observation vectors sampled from an underlying distribution $\mathcal{X}_i \in \RR^{n_i}$, the idea is to find two weight vectors $w_i \in \RR^{n_i}$ and $w_j \in \RR^{n_j}$ so that the random variables $w_i^T \cdot \mathcal{X}_i$ and $w_j^T \cdot \mathcal{X}_j$ are maximally correlated ($w_i$ and $w_j$ are used to map the random vectors to random variables, by computing weighted sums of vector components). Let $\rho(x,y)$ denote the sample-based correlation coefficient between two vectors of observations $x$ and $y$. By using the sample matrix notation $X_i$ and $X_j$ (assuming no data is missing to simplify the exposition), this problem can be formulated as the following optimization problem:
\begin{equation*}
\begin{aligned}
& \underset{w_i \in \RR^{n_i}, w_j \in \RR^{n_j}}{\text{maximize}}
& & \rho(w_i^T X_i , w_j^T X_j) = \frac{w_i^T C_{i,j} w_j}{\sqrt{w_i^T C_{i,i} w_i} \sqrt{w_j^T C_{j,j} w_j}},
\end{aligned}
\end{equation*}
where $C_{i,i}$ and $C_{j,j}$ are empirical estimates of variances of $\mathcal{X}_i$ and $\mathcal{X}_j$ respectively and $C_{i,j}$ is an estimate for the covariance matrix. Assuming that the observation vectors are centered (only for the purposes of presentation), the matrices are computed in the following way: $C_{i,j} = \frac{1}{n-1}X_i X_j^T$, and similarly for $C_{i,i}$ and $C_{j,j}$.
The optimization problem can be reduced to an eigenvalue problem and includes inverting the variance matrices $C_{i,i}$ and $C_{j,j}$. If the matrices are not invertible, one can use a regularization technique by replacing $C_{i,i}$ with $(1- \kappa)C_{i,i} + \kappa I$, where $\kappa \in [0,1]$ is the regularization coefficient and $I$ is the identity matrix. (The same can be applied to $C_{j,j}$.)
A single canonical variable is usually inadequate in representing the original random vector and typically one looks for $k$ projection pairs $(w_i^1, w_j^1),\ldots,(w_i^k, w_j^k)$, so that $(w_i^{u})^T \mathcal{X}_i$ and $(w_j^{u})^T \mathcal{X}_j$ are highly correlated and $(w_i^{u})^T \mathcal{X}_i$ is uncorrelated with $(w_i^{v})^T \mathcal{X}_i$  for $u \neq v$ and analogously for $w_j^u$ vectors.

@@

\subsection{Hub language based CCA Extension}\label{sec:hublang}
Building cross-lingual similarity models based on comparable corpora is challenging for two main reasons. The first problem is related to missing alignment data: when a number of languages is large, the dataset of documents that cover all languages is small (or may even be empty). Even if only two languages are considered, the set of aligned documents can be small (an extreme example is given by the Piedmontese and Hindi Wikipedias where no inter-language links are available), in which case none of the methods presented so far are applicable.
 The second challenge is scale - the data is high-dimensional (many languages with hundreds of thousands of features per language) and the number of multilingual documents may be large (over one million in case of Wikipedia). The optimization problem posed by CCA is not
 trivial to solve: the covariance matrices themselves are prohibitively large to fit in memory (even storing a 100,000 by 100,000 element matrix requires 80GB of memory) and iterative matrix-multiplication based approaches to  solving generalized eigenvalue problems are required (the covariance matrices can be expressed as products of sparse matrices, which means we have fast matrix-vector multiplication).

We now describe an extension of CCA to more than two languages, which can be trained on large comparable corpora and can handle missing data.
 The extension we consider is based on a generalization of CCA to more than two views, introduced in~\cite{Kettenring}, namely the Sum of Squared Correlations SSCOR, which we will state formally later in this section. Our approach exploits a certain characteristic of the data, namely the \emph{hub language} characteristic (see below) in two
 ways: to reduce the dimensionality of the data and to simplify the optimization problem.

>>>>

\jairheading{55}{2016}{x-xx}{3/24}{1/16}
\ShortHeadings{Cross-Lingual Document Similarity and Event Tracking}
{Rupnik, Muhi\v{c}, Leban, Fortuna, \v{S}kraba \& Grobelnik}
\firstpageno{1}

@@

In today's world, we follow news which is distributed globally. Significant events are reported by different sources and in different languages. In this work, we address the problem of tracking of events in a large multilingual stream. Within a recently developed system Event Registry we examine two aspects of this problem: how to compare articles in different languages and how to link collections of articles in different languages which refer to the same event.  Taking a multilingual stream and clusters of articles from each language, we compare different cross-lingual document similarity measures based on Wikipedia. This allows us to compute the similarity of any two articles regardless of language. Building on previous work, we show there are methods which scale well and can compute a meaningful similarity between articles from languages with little or no direct overlap in the training data.
Using this capability, we then propose an approach to link clusters of articles across languages which represent the same event. We provide an extensive evaluation of the system as a whole, as well as an evaluation of the quality and robustness of the similarity measure and the linking algorithm.
\end{abstract}

@@

Content on the Internet is becoming increasingly multilingual. A prime example is Wiki-pedia. In 2001, the majority of pages were written in English, while in 2015, the percentage of English articles has dropped to 14\%. At the same time, online news has begun to dominate reporting of current events. However, machine translation remains relatively rudimentary. It allows people to understand simple phrases on web pages, but remains inadequate for more advanced understanding of text. In this paper we consider the intersection of these developments: how to track events which are reported about in multiple languages.

@@

Collected articles are first semantically annotated by identifying mentions of relevant concepts -- either entities or important keywords. The disambiguation and entity linking of the concepts is done using Wikipedia as the main knowledge base. The algorithm for semantic annotation uses machine learning to detect significant terms within unstructured text and link them to the appropriate Wikipedia articles. The approach models link probability and combines prior word sense distributions with context based sense distributions. The details are reported by
~\citeauthor{Milne:2008:LLW:1458082.1458150}~\citeyear{Milne:2008:LLW:1458082.1458150} and~\citeauthor{zhang2014saaacamactat}~\citeyear{zhang2014saaacamactat}. As a part of the semantic annotation we also analyze the dateline of the article to identify the location of the described event as well as to extract dates mentioned in the article using a set of regular expressions. Since articles are frequently revised we also detect if the collected article is just a revision of a previous one so that we can use this information in next phases of the pipeline. The last important processing step on the document level is to efficiently identify which articles in other available languages are most similar to this article. The methodology for this task is one of the main contributions of this paper and is explained in details in Section~\ref{sec:crosslingual}.

As the next step, an online clustering algorithm~\cite{brank2014} is applied to the articles in order to identify groups of articles that are discussing the same event. For each new article, the clustering algorithm determines if the article should be assigned to some existing cluster or into a new cluster. The underlying assumption is that articles that are describing the same event are similar enough and will therefore be put into the same cluster. For clustering, each new article is first tokenized, stop words are removed and the remaining words are stemmed.
The remaining tokens are represented in a vector-space model and normalized using TF-IDF\footnote{The IDF weights are dynamically computed for each new article over all news articles within a 10 day window.} (see Section~\ref{sec:tfidf} for the definition). Cosine similarity is used to find the most similar existing cluster, by comparing the document's vector to the centroid vector of each cluster. A user-defined threshold is used to determine if the article is not similar enough to any existing clusters (0.4 was used in our experiments).  If the highest similarity is above the threshold, the article is assigned to the corresponding cluster, otherwise a new cluster is created, initially containing only the single article. Whenever an article is assigned to a cluster, the cluster's centroid vector is also updated. Since articles about an event are commonly written only for a short period of time, we remove clusters once the oldest article in the cluster becomes more than 4 days old. This housekeeping mechanism prevents the clustering from becoming slow and also ensures that articles are not assigned to obsolete clusters.

@@

When a cluster is identified and information about the event is extracted, all available data is stored in a custom-built database system. The data is then accessible through the API or a web interface (\url{http://eventregistry.org/}), which provide numerous search and visualization options.

@@

\section{Related Work}\label{sec:related}

@@

\subsection{Cross-Lingual Document Similarity}

@@

\subsubsection{Translation and Dictionary Based Approaches} The most obvious way to compare documents written in different languages is to use machine
translation and perform monolingual similarity \cite<see>{multilingualBook,plagiarism}. One can use free tools such as Moses~\shortcite{moses} or translation services, such as Google Translate (\mbox{\url{https://translate.google.com/}}). There are two issues with such approaches: they solve a harder problem than needs to be solved and they are less robust to training resource quality - large sets of translated sentences are typically needed. Training Moses for languages with scarce linguistic resources is thus problematic. The issue with using online services such as Google Translate is that the APIs are limited and not free. The operation efficiency and cost requirements make translation-based approaches less suited for our system. Closely related are works Cross-Lingual Vector Space Model (CL-VSM)~\cite{plagiarism} and the approach presented by~\citeauthor{pouliquen2008story}~\citeyear{pouliquen2008story} which both compare documents by using dictionaries, which in both cases are EuroVoc dictionaries~\cite{eurovoc}. The generality of such approaches is limited by the quality of available linguistic resources, which may be scarce or non-existent for certain language pairs.

\subsubsection{Probabilistic Topic Model Based Approaches} There exist many variants to modelling documents in a language independent way by using probabilistic graphical models. The models include:  Joint Probabilistic Latent Semantic Analysis (JPLSA)~\shortcite{platt2010translingual}, Coupled Probabilistic LSA (CPLSA)~\cite{platt2010translingual}, Probabilistic Cross-Lingual LSA (PCLLSA)~\cite{PCL_LSA} and Polylingual Topic Models (PLTM)~\shortcite{polyLDA} which is a Bayesian version of PCLLSA. The methods (except for CPLSA) describe the multilingual document collections as samples from generative probabilistic models, with variations on the assumptions on the model structure. The topics represent latent variables that are used to generate observed variables (words), a process specific to each language. The parameter estimation is posed as an inference problem which is typically intractable and one usually solves it using approximate techniques. Most variants of solutions are based on Gibbs sampling or Variational Inference, which are nontrivial to implement and may require an experienced practitioner to be applied. Furthermore, representing a new document as a mixture of topics is another potentially hard inference problem which must be solved.

\subsubsection{Matrix Factorization Base Approaches} Several matrix factorization based approaches exist in the literature. The models include: Non-negative matrix factorization based~\cite{nonnegfactor_lsi}, Cross-Lingual Latent Semantic Indexing CL-LSI~\shortcite{cl_lsi,multilingualBook}, Canonical Correlation Analysis (CCA)~\cite{Hotelling}, Oriented Principal Component Analysis (OPCA)~\cite{platt2010translingual}. The quadratic time and space dependency of the OPCA method makes it impractical for large scale purposes. In addition, OPCA forces the vocabulary sizes for all languages to be the same, which is less intuitive. For our setting, the method in~\cite{nonnegfactor_lsi} has a prohibitively high computational cost when building models (it uses dense matrices whose dimensions are a product of the training set size and the vocabulary size). Our proposed approach combines CCA and CL-LSI. Another closely related method is Cross-Lingual Explicit Semantic Analysis (CL-ESA)~\cite{ESA}, which uses Wikipedia (as do we in the current work) to compare documents. It can be interpreted as using the sample covariance matrix between features of two languages to define the dot product which is used to compute similarities.
The authors of CL-ESA compare it to CL-LSI and find that CL-LSI can outperform CL-ESA in an information retrieval, but is costlier to optimize over a large corpus (CL-ESA requires no training). We find that the scalability argument does not apply in our case: based on advances in numerical linear algebra we can solve large CL-LSI problems that involve millions of documents as opposed to the 10,000 document limit reported by~\citeauthor{ESA}~\citeyear{ESA}. In addition, CL-ESA is less suited for computing similarities between two large monolingual streams. For example, each day we have to compute similarities between 500,000 English and 500,000 German news articles. Comparing each German news article with 500,000 English news articles is either prohibitively slow (involves projecting all English articles on Wikipedia) or consumes too much memory (involves storing the projected English articles, which for a Wikipedia of size 1,000,000 is a 500,000 by 1,000,000 non-sparse matrix).

\subsubsection{Monolingual Approaches} Finally, related work includes monolingual approaches that treat document written in different languages in a monolingual fashion. The intuition is that named entities (for example, ``Obama'') and cognate words (for example, ``tsunami'') are written in the same or similar fashion in many languages. For example, the Cross-Language Character n-Gram Model (CL-CNG)~\cite{plagiarism} represents documents as bags of character $n$-grams. Another approach is to use language dependent keyword lists based on cognate words~\cite{pouliquen2008story}. These approaches may be suitable for comparing documents written in languages that share a writing system, which does not apply to the case of global news tracking.

@@

\subsection{Cross-Lingual Cluster Linking}

@@

\begin{sloppypar}
A system, which is significantly different but worth mentioning, is the GDELT project~\cite{Leetaru2013Gdelt}. In GDELT, events are also extracted from articles, but in their case, an event is specified in a form of a triple containing two actors and a relation. The project contains an extensive vocabulary of possible relations, mostly related to political events. In order to identify events, GDELT collects articles in more than 65 languages and uses machine translation to translate them to English. All information extraction is then done on the translated article.
\end{sloppypar}

\section{Cross-Lingual Document Similarity}
\label{sec:crosslingual}
Document similarity is an important component in techniques from text mining and natural language processing. Many techniques use the similarity as a black box, e.g., a kernel in Support Vector Machines. Comparison of documents (or other types of text snippets) in a monolingual setting is a well-studied problem in the field of information retrieval~\cite{Salton88term-weightingapproaches}. We first formally introduce the problem followed by a description of  our approach.

\subsection{Problem Definition}\label{sec:tfidf}
We will first describe how documents are represented as vectors and how to compare documents in a mono-lingual setting. We then define a way to measure cross-lingual similarity which is natural for the models we consider.

\subsubsection{Document Representation}
The standard vector space model~\cite{Salton88term-weightingapproaches} represents documents as vectors, where each term corresponds to a word or a phrase in a fixed vocabulary. Formally, document $d$ is represented by a vector $x \in \RR^n$, where $n$ corresponds to the size of the vocabulary, and vector elements $x_k$ correspond to the number of times term $k$ occurred in the document, also called \emph{term frequency} or $TF_k(d)$.

@@

\subsubsection{Mono-Lingual Similarity}
A common way of computing similarity between documents is \emph{cosine similarity},
$$sim(d_1, d_2) = \frac{\langle \phi(d_1), \phi(d_2)\rangle}{\|\phi(d_1)\| \|\phi(d_2)\|},$$
where $\langle \cdot,\cdot \rangle$ and $\|\cdot\|$ are standard inner product and Euclidean norm. When dealing with two or more languages, one could ignore the language information
and build a vector space using the union of tokens over the languages. A cosine similarity function in such a space can be useful to some extent, for example ``Internet'' or ``Obama'' may appear both in Spanish and English texts and the presence of such terms in both an English and a Spanish document would contribute to their similarity. In general however, large parts of vocabularies may not intersect. This means that given a language pair, many words in both languages cannot contribute to the similarity score. Such cases can make the similarity function very insensitive to the data.

\subsubsection{Cross-Lingual Similarity}
Processing a multilingual dataset results in several vector spaces with varying dimensionality, one for each language. The dimensionality of the vector space corresponding to the $i$-th language is denoted by $n_i$ and the vector space model mapping is denoted by $\phi_i : \text{text} \rightarrow \RR^{n_i}$.
The similarity between documents in language $i$ and language $j$ is defined as a bilinear operator represented as a matrix $S_{i,j} \in \RR^{n_i \times n_j}$:
$$sim_{i,j}(d_1, d_2) = \frac{ \langle \phi_i (d_1), S_{i,j} \phi_j (d_2) \rangle }{\|\phi_i(d_1)\| \|\phi_j(d_2)\|},$$
where $d_1$ and $d_2$ are documents written in the $i$-th and $j$-th language respectively. If the maximal singular value of $S_{i,j}$ is bounded by $1$, then the similarity scores will lie on the interval $[-1, 1]$. We will provide an overview of the models in Section \ref{sec:models} and then introduce additional notation in \ref{sec:notation}. Starting with Section \ref{sec:kmeans} and ending with Section \ref{sec:hublang} we will describe some approaches to compute $S_{i,j}$ given training data.

\subsection{Cross-Lingual Models}\label{sec:models}
In this section, we will describe several approaches to the problem of computing the multilingual similarities introduced in Section~\ref{sec:tfidf}. We present four approaches:
a simple approach based on $k$-means clustering in Section~\ref{sec:kmeans}, a standard approach based on singular value decomposition in Section~\ref{sec:LSI}, a related
approach called Canonical Correlation Analysis (CCA) in Section~\ref{sec:CCA} and finally a new method, which is an extension of CCA to more than two languages in Section~\ref{sec:hublang}.
%
CCA can be used to find correlated patterns for a pair of languages, whereas the extended method optimizes a
Sum of Squared Correlations (SSCOR) between several language pairs, which was introduced by~\citeauthor{Kettenring}~\citeyear{Kettenring}. The SSCOR problem is difficult to solve in our setting (hundreds of thousands of features, hundreds of thousands of examples). To tackle this, we propose a method which consists of two ingredients.
 The first one is based on an observation that certain datasets (such as Wikipedia) are biased towards one language (English for Wikipedia), which can be exploited
 to reformulate a difficult optimization problem as an eigenvector problem. The second ingredient is dimensionality reduction using CL-LSI, which
 makes the eigenvector problem computationally and numerically tractable.

@@

We will now describe four models to cross-lingual similarity computation in the next sub-sections.
\subsection{$k$-Means}\label{sec:kmeans}

@@

\subsubsection{Implementation Note}

@@

\subsection{Canonical Correlation Analysis}\label{sec:CCA}
 We now present a statistical technique to analyze data from two sources, an extension of which will be presented in the next section.
\begin{sloppypar}
 Canonical Correlation Analysis (CCA)~\cite{Hotelling} is a dimensionality reduction technique similar to Principal Component Analysis (PCA)~\cite{Pearson1901On}, with the additional assumption that the data consists of feature vectors that arose from two sources (two views) that share some information. Examples include: bilingual document collection~\shortcite{mrpqr} and collection of images and captions~\shortcite{Hardoon_usingimage}. Instead of looking for linear combinations of features that maximize the variance (PCA) we look for a linear combination of feature vectors from the first view and a linear combination for the second view, that are maximally correlated.
\end{sloppypar}
Interpreting the columns of $X_i$ as observation vectors sampled from an underlying distribution $\mathcal{X}_i \in \RR^{n_i}$, the idea is to find two weight vectors $w_i \in \RR^{n_i}$ and $w_j \in \RR^{n_j}$ so that the random variables $w_i^T \cdot \mathcal{X}_i$ and $w_j^T \cdot \mathcal{X}_j$ are maximally correlated ($w_i$ and $w_j$ are used to map the random vectors to random variables, by computing weighted sums of vector components). Let $\rho(x,y)$ denote the sample-based correlation coefficient between two vectors of observations $x$ and $y$. By using the sample matrix notation $X_i$ and $X_j$ (assuming no data is missing to simplify the exposition), this problem can be formulated as the following optimization problem:
\begin{equation*}
\begin{aligned}
& \underset{w_i \in \RR^{n_i}, w_j \in \RR^{n_j}}{\text{maximize}}
& & \rho(w_i^T X_i , w_j^T X_j) = \frac{w_i^T C_{i,j} w_j}{\sqrt{w_i^T C_{i,i} w_i} \sqrt{w_j^T C_{j,j} w_j}},
\end{aligned}
\end{equation*}
where $C_{i,i}$ and $C_{j,j}$ are empirical estimates of variances of $\mathcal{X}_i$ and $\mathcal{X}_j$ respectively and $C_{i,j}$ is an estimate for the covariance matrix. Assuming that the observation vectors are centered (only for the purposes of presentation), the matrices are computed in the following way: $C_{i,j} = \frac{1}{n-1}X_i X_j^T$, and similarly for $C_{i,i}$ and $C_{j,j}$.
The optimization problem can be reduced to an eigenvalue problem and includes inverting the variance matrices $C_{i,i}$ and $C_{j,j}$. If the matrices are not invertible, one can use a regularization technique by replacing $C_{i,i}$ with $(1- \kappa)C_{i,i} + \kappa I$, where $\kappa \in [0,1]$ is the regularization coefficient and $I$ is the identity matrix. (The same can be applied to $C_{j,j}$.)
A single canonical variable is usually inadequate in representing the original random vector and typically one looks for $k$ projection pairs $(w_i^1, w_j^1),\ldots,(w_i^k, w_j^k)$, so that $(w_i^{u})^T \mathcal{X}_i$ and $(w_j^{u})^T \mathcal{X}_j$ are highly correlated and $(w_i^{u})^T \mathcal{X}_i$ is uncorrelated with $(w_i^{v})^T \mathcal{X}_i$  for $u \neq v$ and analogously for $w_j^u$ vectors.

@@

\subsection{Hub Language Based CCA Extension}\label{sec:hublang}
Building cross-lingual similarity models based on comparable corpora is challenging for two main reasons. The first problem is related to missing alignment data: when a number of languages is large, the dataset of documents that cover all languages is small (or may even be empty). Even if only two languages are considered, the set of aligned documents can be small (an extreme example is given by the Piedmontese and Hindi Wikipedias where no inter-language links are available), in which case none of the methods presented so far are applicable.
 The second challenge is scale - the data is high-dimensional (many languages with hundreds of thousands of features per language) and the number of multilingual documents may be large (over one million in case of Wikipedia). The optimization problem posed by CCA is not
 trivial to solve: the covariance matrices themselves are prohibitively large to fit in memory (even storing a 100,000 by 100,000 element matrix requires 80GB of memory) and iterative matrix-multiplication based approaches to  solving generalized eigenvalue problems are required (the covariance matrices can be expressed as products of sparse matrices, which means we have fast matrix-vector multiplication).

We now describe an extension of CCA to more than two languages, which can be trained on large comparable corpora and can handle missing data.
 The extension we consider is based on a generalization of CCA to more than two views, introduced by~\citeauthor{Kettenring}~\citeyear{Kettenring}, namely the Sum of Squared Correlations SSCOR, which we will state formally later in this section. Our approach exploits a certain characteristic of the data, namely the \emph{hub language} characteristic (see below) in two
 ways: to reduce the dimensionality of the data and to simplify the optimization problem.
